### 2016-03-23 doc/draft.Rmd
work on dispersion

**Validation and quality controls in sequence based studies of intrahost diversity**

@@

In these populations with lower diversity and input titer, we maintained greater than 0.85 sensitivity for SNV at 1% frequency or higher. . Despite a high depth of coverage (>10,000 reads per bp), our sensitivity was considerably lower for variants at or below 0.5% frequency (`r Hiseq`B and C). Therefore, the drop in sensitivity is most likely due to the 1,000-fold decrease in nucleic acid concentration compared to the first data set, and the fact that library preparation requires a number of sampling steps that may limit detection. During library preparation, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.

The specificity was greater than 0.99 in all the samples with 10^5^ genomes per microliter. As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome. The false positives outnumber the true positives by 100-fold in these populations with realistic diversity and input. We were able to increased our specificity  by applying a more stringent p-value cutoff. However, as shown by the ROC curves in `r Hiseq`, this move towards the y axis markedly reduces sensitivity. Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.

@@

We have shown that robust benchmarking is vital to accurate NGS based studies of viral diversity, and that differences in experimental design and sample preparation can lead to wide variability in the accuracy of SNV indentification. We found input nucleic acid concentrations, which can vary greatly in patient-derived samples, had large effects on both sensitivity and specificity. At moderate levels of nucleic acid input we could improve accuracy by filtering putative SNV based on quality metrics and read positions. We further improved our accuracy at low input levels by controlling for RT-PCR errors by processing samples in duplicates. While our quality cut offs may not be needed by all variant callers, our work with Lofreq indicates that experimental design is important in accurate SNV detection. This is important given that few variant callers (if any) have been benchmarked under patient-derived conditions. Finally, we show that these inaccuracies in SNV calling drastically impact down stream analysis, and that accurate conclusions require quantifying and limiting this error.

 To our knowledge, DeepSNV is the only variant calling method that has been benchmarked on real sequencing reads in which all true positive variants and their frequencies were known *a priori* and independent of NGS (cite Gerstung). Given this robust benchmarking we thought it ideal for identifying rare variants in populations of influenza virus. However, we found decreased sensitivity and specificity when applied to cell culture derived mixtures of WSN33 and PR8. The most striking drop in sensitivity came for variants found at 0.16%. Under these conditions we found true variants at a rate of `r miseq.sense.16` compared the expected sensitivity of  0.860 for variants at 0.1%. This decrease is most likely due to the fact that we are looking for variants across `r miseq_kb`bp of the influenza genome where previous benchmarking focused on just 1,500 bp of the HIV pol gene. Therefore, our sequencing efforts were diluted across the influenza genome resulting in decreased read depth  which limits sensitivity. Additionally,  we aligned all reads to a PR8 reference genome which biases our detection of WSN33 variants especially rare variants, which already have few reads. We also observed a relatively modest decrease in specificity. Despite the use of bonferronni corrected p values, the drop in specificity is most likely due to multiple testing errors caused by applying the algorithm across `r miseq_kb`bp (cite).

@@

 iv) could duplicates have been pooled and sequenced as one sample?

IV. Comparision to Lofreq  and Consequences of errors
  i) our benchmarking shows that DeepSNVs means of distinguishing true from false variants is partitially breaks down under conditions that mimic patient derived samples.
  ii) This can be overcome with the addition of specific quality filters.
  iii) We would not have known about these inaccuracies had we not run our benchmarking experiment.
  iv) Do other variant callers have similar inaccuracies.
  v) Yes Lofreq, although more accurate than our unmodified DeepSNV pipeline, had decreased specificity than had previously been reported.
  vi) The initial Lofreq sensitivity and specificity metrics were based largely on a *in silico* data sets.
  vii) Lofreq did not display this high accuracy in under our experimental conditions.
  viii) While we did not robustly attempt to improve the Lofreq accuracy we did find that frequency threshold could improve its specificity.
  ix) Experimental conditions matter, regardless of variant calling method and so accurate study requires controls to be run.

>>>>

**Validation and quality controls in next generation sequencing based studies of intrahost diversity**

@@

In these populations with lower diversity and input titer, we maintained greater than 0.85 sensitivity for SNV at 1% frequency or higher. Despite a high depth of coverage (>10,000 reads per bp), our sensitivity was considerably lower for variants at or below 0.5% frequency (`r Hiseq`B and C). The drop in sensitivity is most likely due to the 1,000-fold decrease in nucleic acid concentration compared to the first data set, and the fact that library preparation requires a number of sampling steps that may limit detection. During library preparation, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.

In our initial analysis of these data using the same default DeepSNV settings as above, the specificity was significantly lower, 0.xxx. These lower input samples underwent more PCR cycles, which skews the noise distribution of variant calls in the test vs. plasmid control samples. Therefore, we utilized an alternative beta binomial model available in DeepSNV, which more appropriately models these differences (cite). With these settings, the specificity was greater than 0.99 in all the samples with 10^5^ genomes per microliter. As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome. The false positives outnumber the true positives by 100-fold in these populations with realistic diversity and input. We were able to increased our specificity  by applying a more stringent p-value cutoff. However, as shown by the ROC curves in `r Hiseq`, this move towards the y axis markedly reduces sensitivity. Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.

@@

Robust validation is essential in NGS-based studies of viral diversity. Differences in experimental design and sample preparation can lead to wide variability in the accuracy of SNV indentification. We found that input nucleic acid concentration, which can vary greatly in patient-derived samples, had a large impact on both the sensitivity and specificity of rare variant detection. At moderate levels of nucleic acid input we could improve accuracy by filtering putative SNV based on quality metrics and read position. We further improved our accuracy at low input levels by processing these samples in duplicate. While our quality cut-offs may not be universally applicable to all samples and variant callers, our data sugest that experimental design is critical for accurate SNV detection. These findings are important as few, if any, variant callers have been benchmarked under patient-derived conditions. Finally, we show that these inaccuracies in SNV calling drastically impact downstream analysis.

We chose DeepSNV for our studies, because it is one of the only variant callers that has been validated on real sequencing reads in which all true positive variants and their frequencies were known *a priori* and independent of NGS (cite Gerstung). In our experimental conditions we found decreased sensitivity and specificity when applied to cell culture derived mixtures of WSN33 and PR8. The most striking drop in sensitivity came for variants found at 0.16%. Under these conditions we found true variants at a rate of `r miseq.sense.16` compared the expected sensitivity of  0.860 for variants at 0.1%. This decrease is most likely due to the fact that we are looking for variants across `r miseq_kb`bp of the influenza genome where previous benchmarking focused on just 1,500 bp of the HIV pol gene. Therefore, our sequencing efforts were diluted across the influenza genome resulting in decreased read depth  which limits sensitivity. Additionally,  we aligned all reads to a PR8 reference genome which biases our detection of WSN33 variants especially rare variants, which already have few reads. We also observed a relatively modest decrease in specificity. Despite the use of bonferronni corrected p values, the drop in specificity is most likely due to multiple testing errors caused by applying the algorithm across `r miseq_kb`bp (cite).

@@
